---
<<<<<<< HEAD
title: "Intersectionality using Iterative Proportional Fitting on 2021 UK Census Data"
author: "Richard Blackwell"
date: 2026-08-07
format: 
  html:
    toc: true
    toc-location: left
    toc_depth: 3
    toc_float: true
    toc-expand: true
    embed-resources: true
    css: hisw.css
    grid:
      sidebar-width: 200px
      body-width: 1600px
      margin-width: 50px
      gutter-width: 1.5rem
=======
title: "Intersectionality using Iterative Proportional Fitting of 2021 UK Census Data"
author: "Richard Blackwell"
format: html
editor: visual
>>>>>>> 2da3d9fbd26f11c30322a8fd54f58c56dcc63303
---

```{r}
#| label: init
#| echo: false
#| warning: false

library(tidyverse)
library(flextable)
library(officer)
<<<<<<< HEAD
```

## Iterative Proportional Fitting - A simple worked example

```{r}
#| label: tbl-area
#| tbl-cap-location: bottom
#| tbl-cap: "Populations by Area"
#| echo: false
#| warning: false

df_area <- data.frame(AREA = c("A", "B", "C"),
                        POPN = c(80, 70, 84))

tbl_area <- flextable(df_area) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    POPN = "Popn.") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header")
tbl_area
```

## Marginal Data

### Gender

##### Counts

```{r}
#| label: tbl-gender
#| tbl-cap-location: bottom
#| tbl-cap: "Gender Marginals - Counts"
#| echo: false
#| warning: false

df_gender <- data.frame(AREA = c("A", "B", "C"),
                        FEMALE = c(45, 40, 49),
                        MALE = c(35, 30, 35))

tbl_gender <- flextable(df_gender %>% 
                          bind_rows(df_gender %>% summarise(across(2:3, sum)) %>% mutate(AREA = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:3]))) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    FEMALE = "Female", 
                    MALE = "Male",
                    TOTAL = "Total") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_gender), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_gender), border = fp_border(width = 1), part = "body")
tbl_gender
```

##### Percentages

```{r}
#| label: tbl-gender-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Gender Marginals - Percentages"
#| echo: false
#| warning: false

tbl_gender_pct <- flextable(df_gender %>% 
                              mutate(TOTAL = rowSums(.[2:3])) %>%
                              mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                              select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    FEMALE = "Female", 
                    MALE = "Male") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:3, digits = 1, suffix = "%")
tbl_gender_pct
```

### Age

##### Counts

```{r}
#| label: tbl-age
#| tbl-cap-location: bottom
#| tbl-cap: "Age Marginals - Counts"
#| echo: false
#| warning: false

df_age <- data.frame(AREA = c("A", "B", "C"),
                     YOUNG = c(20, 25, 31),
                     ADULT = c(30, 25, 33),
                     ELDERLY = c(30, 20, 20))

tbl_age <- flextable(df_age %>%
                       bind_rows(df_age %>% summarise(across(2:4, sum)) %>% mutate(AREA = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:4]))) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    YOUNG = "Young", 
                    ADULT = "Adult",
                    ELDERLY = "Elderly",
                    TOTAL = "Total") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_age), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_age), border = fp_border(width = 1), part = "body")
tbl_age
```

##### Percentages

```{r}
#| label: tbl-age-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Age Marginals - Percentages"
#| echo: false
#| warning: false

tbl_age_pct <- flextable(df_age %>% 
                           mutate(TOTAL = rowSums(.[2:4])) %>%
                           mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                           select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    YOUNG = "Young", 
                    ADULT = "Adult",
                    ELDERLY = "Elderly") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:4, digits = 1, suffix = "%")
tbl_age_pct
```

### Health

##### Counts

```{r}
#| label: tbl-health
#| tbl-cap-location: bottom
#| tbl-cap: "Health Marginals - Counts"
#| echo: false
#| warning: false

df_health <- data.frame(AREA = c("A", "B", "C"),
                        GOOD = c(16, 14, 18),
                        FAIR = c(30, 28, 33),
                        POOR = c(34, 28, 33))

tbl_health <- flextable(df_health %>%
                       bind_rows(df_health %>% summarise(across(2:4, sum)) %>% mutate(AREA = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:4]))) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor",
                    TOTAL = "Total") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_health), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_health), border = fp_border(width = 1), part = "body")
tbl_health
```

##### Percentages

```{r}
#| label: tbl-health-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Health Marginals - Percentages"
#| echo: false
#| warning: false

tbl_health_pct <- flextable(df_health %>% 
                              mutate(TOTAL = rowSums(.[2:4])) %>%
                           mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                           select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:4, digits = 1, suffix = "%")
tbl_health_pct
```

## Constraints

### Gender x Age

##### Counts

```{r}
#| label: tbl-gender-age
#| tbl-cap-location: bottom
#| tbl-cap: "Gender x Age Constraints - Counts"
#| echo: false
#| warning: false

df_gender_age <- data.frame(GENDER = c("Female", "Male"),
                            YOUNG = c(36, 40),
                            ADULT = c(59, 29),
                            ELDERLY = c(39, 31))

tbl_gender_age <- flextable(df_gender_age %>%
                              bind_rows(df_gender_age %>% summarise(across(2:4, sum)) %>% mutate(GENDER = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:4]))) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Gender", "Age", "Total"), colwidths = c(1, 3, 1)) %>%
  set_header_labels(GENDER = "Gender",
                    YOUNG = "Young", 
                    ADULT = "Adult",
                    ELDERLY = "Elderly",
                    TOTAL = "Total") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_gender_age), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_gender_age), border = fp_border(width = 1), part = "body")
tbl_gender_age
```



##### Percentages

```{r}
#| label: tbl-gender-age-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Gender x Age Constraints - Percentages"
#| echo: false
#| warning: false

tbl_gender_age_pct <- flextable(df_gender_age %>% 
                                  mutate(TOTAL = rowSums(.[2:4])) %>%
                                  mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                                  select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Gender", "Age"), colwidths = c(1, 3)) %>%
  set_header_labels(GENDER = "Gender",
                    YOUNG = "Young", 
                    ADULT = "Adult",
                    ELDERLY = "Elderly") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:4, digits = 1, suffix = "%")
tbl_gender_age_pct
```


### Gender x Health

##### Counts

```{r}
#| label: tbl-gender-health
#| tbl-cap-location: bottom
#| tbl-cap: "Gender x Health Constraints - Counts"
#| echo: false
#| warning: false

df_gender_health <- data.frame(GENDER = c("Female", "Male"),
                               GOOD = c(25, 23),
                               FAIR = c(53, 38),
                               POOR = c(56, 39))

tbl_gender_health <- flextable(df_gender_health %>%
                                 bind_rows(df_gender_health %>% summarise(across(2:4, sum)) %>% mutate(GENDER = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:4]))) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Gender", "Health", "Total"), colwidths = c(1, 3, 1)) %>%
  set_header_labels(GENDER = "Gender",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor",
                    TOTAL = "Total") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_gender_health), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_gender_health), border = fp_border(width = 1), part = "body")
tbl_gender_health
```



##### Percentages

```{r}
#| label: tbl-gender-health-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Gender x Health Constraints - Percentages"
#| echo: false
#| warning: false

tbl_gender_health_pct <- flextable(df_gender_health %>% 
                                     mutate(TOTAL = rowSums(.[2:4])) %>%
                                     mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                                     select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Gender", "Health"), colwidths = c(1, 3)) %>%
  set_header_labels(GENDER = "Gender",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:4, digits = 1, suffix = "%")
tbl_gender_health_pct
```

### Age x Health

##### Counts

```{r}
#| label: tbl-age-health
#| tbl-cap-location: bottom
#| tbl-cap: "Age x Health Constraints - Counts"
#| echo: false
#| warning: false

df_age_health <- data.frame(AGE = c("Young", "Adult", "Elderly"),
                            GOOD = c(12, 16, 20),
                            FAIR = c(35, 31, 25),
                            POOR = c(29, 41, 25))

tbl_age_health <- flextable(df_age_health %>%
                              bind_rows(df_age_health %>% summarise(across(2:4, sum)) %>% mutate(AGE = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:4]))) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Age", "Health", "Total"), colwidths = c(1, 3, 1)) %>%
  set_header_labels(AGE = "Age",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor",
                    TOTAL = "Total") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_age_health), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_age_health), border = fp_border(width = 1), part = "body")
tbl_age_health
```

##### Percentages

```{r}
#| label: tbl-age-health-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Age x Health Constraints - Percentages"
#| echo: false
#| warning: false

tbl_age_health_pct <- flextable(df_age_health %>% 
                                  mutate(TOTAL = rowSums(.[2:4])) %>%
                                  mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                                  select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Age", "Health"), colwidths = c(1, 3)) %>%
  set_header_labels(AGE = "Age",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:4, digits = 1, suffix = "%")
tbl_age_health_pct
```

### Marginal Data 2021 UK Census

The following list outlines the inclusion domains and underlying marginal variables from the 2021 UK Census and the number of categories for the domain or variable.

-   Communication (31)
    -   Age (8)
    -   Highest Level of Qualification (7)
    -   Main Language (11)
    -   Proficiency in English Language (5)
-   Employment Status (32)
    -   Highest Level of Qualification (7)
    -   NS-SeC (10)
    -   Occupation (current) (10)
    -   Tenure of Household (5)
-   Physical and Practical Access (18)
    -   Age (8)
    -   Car Availability (3)
    -   Disability (3)
    -   General Health (4)
-   Social and Community Connectedness (28)
    -   Car Availability (3)
    -   Ethnicity (6)
    -   Household Composition (8)
    -   Length of Residence in UK (6)
    -   Religion (5)
-   Structural Inclusion (23)
    -   Age (8)
    -   Disability (3)
    -   Ethnicity (6)
    -   General Health (4)
    -   Sex (2)
-   Timing and Caring Commitments (29)
    -   Economic Activity Status (10)
    -   Household Composition (8)
    -   Household Type (6)
    -   Unpaid Care (5)
=======

```

## Worked Example

### Marginals

```{r}
#| label: marginal_popn
#| echo: false
#| warning: false

df_popn <- data.frame(area = c("A","B","C"),
                      popn = c(293, 222, 172))

flextable(df_popn) %>%
  autofit() %>%
  set_header_labels(area = "Area", popn = "Popn") %>%
  vline(j = 1, border = fp_border(width = 2))
```

```{r}
#| label: marginal_age
#| echo: false
#| warning: false

df_age <- data.frame(age = rep(c("Young", "Adult", "Elderly"), 3),
                     area = c(rep("A", 3), rep("B", 3), rep("C", 3)),
                     popn = c(c(90, 133, 70), c(62, 107, 53), c(45, 83, 44)))
df_age <- df_age %>% group_by(area) %>% mutate(pct = popn/sum(popn)) %>% ungroup() %>% select(-popn)

flextable(df_age %>% pivot_wider(names_from = "area", values_from = "pct")) %>%
  autofit() %>%
  align(j = 2:4, align = "center", part = "header") %>%
  add_header_row(values = c("Age", "Area"), colwidths = c(1, 3)) %>%
  set_header_labels(age = "Age") %>%
  merge_at(i = 1:2, j = 1, part = "header") %>%
  vline(j = 1, border = fp_border(width = 2)) %>%
  colformat_double(j = 1:length(unique(df_age$area))+1, digits=3)
```

```{r}
#| label: marginal_gender
#| echo: false
#| warning: false

df_gender <- data.frame(gender = rep(c("Female", "Male"), 3),
                        area = c(rep("A", 2), rep("B", 2), rep("C", 2)),
                        popn = c(c(153, 140), c(116, 106), c(90, 82)))

df_gender <- df_gender %>% group_by(area) %>% mutate(pct = popn/sum(popn)) %>% ungroup() %>% select(-popn)

flextable(df_gender %>% pivot_wider(names_from = "area", values_from = "pct")) %>%
  autofit() %>%
  align(j = 2:4, align = "center", part = "header") %>%
  add_header_row(values = c("Gender", "Area"), colwidths = c(1, 3)) %>%
  set_header_labels(age = "Gender") %>%
  merge_at(i = 1:2, j = 1, part = "header") %>%
  vline(j = 1, border = fp_border(width = 2)) %>%
  colformat_double(j = 1:length(unique(df_gender$area))+1, digits=3)
```

```{r}
#| label: marginal_health
#| echo: false
#| warning: false

df_health <- data.frame(health = rep(c("Good", "Fair", "Poor"), 3),
                        area = c(rep("A", 3), rep("B", 3), rep("C", 3)),
                        popn = c(c(168, 82, 43), c(127, 63, 32), c(96, 51, 25)))

df_health <- df_health %>% group_by(area) %>% mutate(pct = popn/sum(popn)) %>% ungroup() %>% select(-popn)

flextable(df_health %>% pivot_wider(names_from = "area", values_from = "pct")) %>%
  autofit() %>%
  align(j = 2:4, align = "center", part = "header") %>%
  add_header_row(values = c("Health", "Area"), colwidths = c(1, 3)) %>%
  set_header_labels(age = "Health") %>%
  merge_at(i = 1:2, j = 1, part = "header") %>%
  vline(j = 1, border = fp_border(width = 2)) %>%
  colformat_double(j = 1:length(unique(df_health$area))+1, digits=3)
```

### Constraints

```{r}
#| label: constraint_age_gender
#| echo: false
#| warning: false

df_age_gender <- data.frame(age = rep(c("Young", "Adult", "Elderly"), 2),
                            gender = c(rep("Female", 3), rep("Male", 3)),
                            popn = c(c(103, 163, 93), c(94, 160, 74)))
df_age_gender <- df_age_gender %>% mutate(pct = popn/sum(popn)) %>% select(-popn)
```

```{r}
#| label: constraint_age_health
#| echo: false
#| warning: false

df_age_health <- data.frame(age = rep(c("Young", "Adult", "Elderly"), 3),
                            health = c(rep("Good", 3), rep("Fair", 3), rep("Poor", 3)),
                            popn = c(c(120, 187, 84), c(48, 94, 54), c(29, 42, 29)))
df_age_health <- df_age_health %>% mutate(pct = popn/sum(popn)) %>% select(-popn)
```

```{r}
#| label: constraint_gender_health
#| echo: false
#| warning: false

df_gender_health <- data.frame(gender = rep(c("Female", "Male"), 3),
                               health = c(rep("Good", 2), rep("Fair", 2), rep("Poor", 2)),
                               popn = c(c(197, 194), c(104, 92), c(58, 42)))
df_gender_health <- df_gender_health %>% mutate(pct = popn/sum(popn)) %>% select(-popn)
```

### Seed Table

The starting seed table $X^{(0)}$ is given by the combined independent probability multiplied by the population to return the initial estimate of people with the combined values.

For example the seed estimate $X^{(0)}_{a,g,h}$ for young, male in good health is given by $$ \begin{align} X^{(0)}_{Y,M,G} &= P(Y)\times{P(M)}\times{P(G)} \\ &= 0.307\times{0.478}\times{0.573} \\ &= 0.084 \end{align} $$

Each cell of the seed table is calculated in the same manner to cover the $ age\times{gender}\times{health} = 3\times{2}\times{3} = 18 $ combinations resulting in the seed table as follows

```{r}
#| label: seed_table
#| echo: false
#| warning: false

area_list <- unique(df_age$area)
age_list <- unique(df_age$age)
gender_list <- unique(df_gender$gender)
health_list <- unique(df_health$health)

df_seed_table <- data.frame(area = as.character(),
                            age = as.character(),
                            gender = as.character(),
                            health = as.character(),
                            pct = as.numeric())

# seed_table <- array(dim = c(length(area_list), 
#               length(age_list),
#               length(gender_list),
#               length(health_list)),
#               dimnames = list(area_list, age_list, gender_list, health_list))

for(area in area_list){
  for(age in age_list){
    for(gender in gender_list){
      for(health in health_list){
        pct <- df_age$pct[df_age$area==area & df_age$age==age] *
          df_gender$pct[df_gender$area==area & df_gender$gender==gender] *
          df_health$pct[df_health$area==area & df_health$health==health]
        df_seed_table <- df_seed_table %>% bind_rows(data.frame(area, age, gender, health, pct))
      }
    }
  }
}

flextable(df_seed_table) %>%
  set_header_labels(area = "Area",
                    age = "Age",
                    gender = "Gender",
                    health = "Health",
                    pct = "Proportion") %>%
  bold(i = 1, part = "header") %>%
  border_inner(border = fp_border(width = 1)) %>%
  merge_v() %>%
  set_formatter(pct = function(x){sprintf("%.3f", x)})
```

### Create calculated constraints

```{r} 
#| label: calculated_constraint_age_gender
#| echo: false
#| warning: false

df_seed_table %>%
  group_by(age, gender) %>% 
  summarise(pct = sum(pct), .groups = "keep")

sum(df_seed_table$pct)
```
>>>>>>> 2da3d9fbd26f11c30322a8fd54f58c56dcc63303
