---
title: "Intersectionality using Iterative Proportional Fitting on 2021 UK Census Data"
author: "Richard Blackwell"
date: 2026-08-07
format: 
  html:
    toc: true
    toc-location: left
    toc_depth: 3
    toc_float: true
    toc-expand: true
    embed-resources: true
    css: hisw.css
    grid:
      sidebar-width: 200px
      body-width: 1600px
      margin-width: 50px
      gutter-width: 1.5rem
---

```{r}
#| label: init
#| echo: false
#| warning: false

library(tidyverse)
library(flextable)
library(officer)
```

## Iterative Proportional Fitting - A simple worked example

```{r}
#| label: tbl-area
#| tbl-cap-location: bottom
#| tbl-cap: "Populations by Area"
#| echo: false
#| warning: false

df_area <- data.frame(AREA = c("A", "B", "C"),
                        POPN = c(80, 70, 84))

tbl_area <- flextable(df_area) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    POPN = "Popn.") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header")
tbl_area
```

## Marginal Data

### Gender

##### Counts

```{r}
#| label: tbl-gender
#| tbl-cap-location: bottom
#| tbl-cap: "Gender Marginals - Counts"
#| echo: false
#| warning: false

df_gender <- data.frame(AREA = c("A", "B", "C"),
                        FEMALE = c(45, 40, 49),
                        MALE = c(35, 30, 35))

tbl_gender <- flextable(df_gender %>% 
                          bind_rows(df_gender %>% summarise(across(2:3, sum)) %>% mutate(AREA = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:3]))) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    FEMALE = "Female", 
                    MALE = "Male",
                    TOTAL = "Total") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_gender), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_gender), border = fp_border(width = 1), part = "body")
tbl_gender
```

##### Percentages

```{r}
#| label: tbl-gender-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Gender Marginals - Percentages"
#| echo: false
#| warning: false

tbl_gender_pct <- flextable(df_gender %>% 
                              mutate(TOTAL = rowSums(.[2:3])) %>%
                              mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                              select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    FEMALE = "Female", 
                    MALE = "Male") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:3, digits = 1, suffix = "%")
tbl_gender_pct
```

### Age

##### Counts

```{r}
#| label: tbl-age
#| tbl-cap-location: bottom
#| tbl-cap: "Age Marginals - Counts"
#| echo: false
#| warning: false

df_age <- data.frame(AREA = c("A", "B", "C"),
                     YOUNG = c(20, 25, 31),
                     ADULT = c(30, 25, 33),
                     ELDERLY = c(30, 20, 20))

tbl_age <- flextable(df_age %>%
                       bind_rows(df_age %>% summarise(across(2:4, sum)) %>% mutate(AREA = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:4]))) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    YOUNG = "Young", 
                    ADULT = "Adult",
                    ELDERLY = "Elderly",
                    TOTAL = "Total") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_age), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_age), border = fp_border(width = 1), part = "body")
tbl_age
```

##### Percentages

```{r}
#| label: tbl-age-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Age Marginals - Percentages"
#| echo: false
#| warning: false

tbl_age_pct <- flextable(df_age %>% 
                           mutate(TOTAL = rowSums(.[2:4])) %>%
                           mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                           select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    YOUNG = "Young", 
                    ADULT = "Adult",
                    ELDERLY = "Elderly") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:4, digits = 1, suffix = "%")
tbl_age_pct
```

### Health

##### Counts

```{r}
#| label: tbl-health
#| tbl-cap-location: bottom
#| tbl-cap: "Health Marginals - Counts"
#| echo: false
#| warning: false

df_health <- data.frame(AREA = c("A", "B", "C"),
                        GOOD = c(16, 14, 18),
                        FAIR = c(30, 28, 33),
                        POOR = c(34, 28, 33))

tbl_health <- flextable(df_health %>%
                       bind_rows(df_health %>% summarise(across(2:4, sum)) %>% mutate(AREA = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:4]))) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor",
                    TOTAL = "Total") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_health), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_health), border = fp_border(width = 1), part = "body")
tbl_health
```

##### Percentages

```{r}
#| label: tbl-health-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Health Marginals - Percentages"
#| echo: false
#| warning: false

tbl_health_pct <- flextable(df_health %>% 
                              mutate(TOTAL = rowSums(.[2:4])) %>%
                           mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                           select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  set_header_labels(AREA = "Area",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:4, digits = 1, suffix = "%")
tbl_health_pct
```

## Constraints

### Gender x Age

##### Counts

```{r}
#| label: tbl-gender-age
#| tbl-cap-location: bottom
#| tbl-cap: "Gender x Age Constraints - Counts"
#| echo: false
#| warning: false

df_gender_age <- data.frame(GENDER = c("Female", "Male"),
                            YOUNG = c(36, 40),
                            ADULT = c(59, 29),
                            ELDERLY = c(39, 31))

tbl_gender_age <- flextable(df_gender_age %>%
                              bind_rows(df_gender_age %>% summarise(across(2:4, sum)) %>% mutate(GENDER = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:4]))) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Gender", "Age", "Total"), colwidths = c(1, 3, 1)) %>%
  set_header_labels(GENDER = "Gender",
                    YOUNG = "Young", 
                    ADULT = "Adult",
                    ELDERLY = "Elderly",
                    TOTAL = "Total") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_gender_age), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_gender_age), border = fp_border(width = 1), part = "body")
tbl_gender_age
```



##### Percentages

```{r}
#| label: tbl-gender-age-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Gender x Age Constraints - Percentages"
#| echo: false
#| warning: false

tbl_gender_age_pct <- flextable(df_gender_age %>% 
                                  mutate(TOTAL = rowSums(.[2:4])) %>%
                                  mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                                  select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Gender", "Age"), colwidths = c(1, 3)) %>%
  set_header_labels(GENDER = "Gender",
                    YOUNG = "Young", 
                    ADULT = "Adult",
                    ELDERLY = "Elderly") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:4, digits = 1, suffix = "%")
tbl_gender_age_pct
```


### Gender x Health

##### Counts

```{r}
#| label: tbl-gender-health
#| tbl-cap-location: bottom
#| tbl-cap: "Gender x Health Constraints - Counts"
#| echo: false
#| warning: false

df_gender_health <- data.frame(GENDER = c("Female", "Male"),
                               GOOD = c(25, 23),
                               FAIR = c(53, 38),
                               POOR = c(56, 39))

tbl_gender_health <- flextable(df_gender_health %>%
                                 bind_rows(df_gender_health %>% summarise(across(2:4, sum)) %>% mutate(GENDER = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:4]))) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Gender", "Health", "Total"), colwidths = c(1, 3, 1)) %>%
  set_header_labels(GENDER = "Gender",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor",
                    TOTAL = "Total") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_gender_health), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_gender_health), border = fp_border(width = 1), part = "body")
tbl_gender_health
```



##### Percentages

```{r}
#| label: tbl-gender-health-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Gender x Health Constraints - Percentages"
#| echo: false
#| warning: false

tbl_gender_health_pct <- flextable(df_gender_health %>% 
                                     mutate(TOTAL = rowSums(.[2:4])) %>%
                                     mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                                     select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Gender", "Health"), colwidths = c(1, 3)) %>%
  set_header_labels(GENDER = "Gender",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:4, digits = 1, suffix = "%")
tbl_gender_health_pct
```

### Age x Health

##### Counts

```{r}
#| label: tbl-age-health
#| tbl-cap-location: bottom
#| tbl-cap: "Age x Health Constraints - Counts"
#| echo: false
#| warning: false

df_age_health <- data.frame(AGE = c("Young", "Adult", "Elderly"),
                            GOOD = c(12, 16, 20),
                            FAIR = c(35, 31, 25),
                            POOR = c(29, 41, 25))

tbl_age_health <- flextable(df_age_health %>%
                              bind_rows(df_age_health %>% summarise(across(2:4, sum)) %>% mutate(AGE = "Total")) %>% 
                          mutate(TOTAL = rowSums(.[2:4]))) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Age", "Health", "Total"), colwidths = c(1, 3, 1)) %>%
  set_header_labels(AGE = "Age",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor",
                    TOTAL = "Total") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  vline(j = NCOL(df_age_health), border = fp_border(width = 1), part = "all") %>%
  hline(i = NROW(df_age_health), border = fp_border(width = 1), part = "body")
tbl_age_health
```

##### Percentages

```{r}
#| label: tbl-age-health-pct
#| tbl-cap-location: bottom
#| tbl-cap: "Age x Health Constraints - Percentages"
#| echo: false
#| warning: false

tbl_age_health_pct <- flextable(df_age_health %>% 
                                  mutate(TOTAL = rowSums(.[2:4])) %>%
                                  mutate(across(2:4, function(x){x*100/TOTAL})) %>%
                                  select(-TOTAL)) %>%
  autofit() %>%
  border_remove() %>%
  add_header_row(values = c("Age", "Health"), colwidths = c(1, 3)) %>%
  set_header_labels(AGE = "Age",
                    GOOD = "Good", 
                    FAIR = "Fair",
                    POOR = "Poor") %>%
  merge_v(part = "header") %>%
  valign(part = "header", valign = "bottom") %>%
  align(i = 1, j = 2:4, part = "header", align = "center") %>%
  #bold(i = 1, part = "header") %>% 
  vline(j = 1, border = fp_border(width = 1), part = "all") %>%
  hline(i = 1:2, border = fp_border(width = 1), part = "header") %>%
  colformat_double(j = 2:4, digits = 1, suffix = "%")
tbl_age_health_pct
```

### Marginal Data 2021 UK Census

The following list outlines the inclusion domains and underlying marginal variables from the 2021 UK Census and the number of categories for the domain or variable.

-   Communication (31)
    -   Age (8)
    -   Highest Level of Qualification (7)
    -   Main Language (11)
    -   Proficiency in English Language (5)
-   Employment Status (32)
    -   Highest Level of Qualification (7)
    -   NS-SeC (10)
    -   Occupation (current) (10)
    -   Tenure of Household (5)
-   Physical and Practical Access (18)
    -   Age (8)
    -   Car Availability (3)
    -   Disability (3)
    -   General Health (4)
-   Social and Community Connectedness (28)
    -   Car Availability (3)
    -   Ethnicity (6)
    -   Household Composition (8)
    -   Length of Residence in UK (6)
    -   Religion (5)
-   Structural Inclusion (23)
    -   Age (8)
    -   Disability (3)
    -   Ethnicity (6)
    -   General Health (4)
    -   Sex (2)
-   Timing and Caring Commitments (29)
    -   Economic Activity Status (10)
    -   Household Composition (8)
    -   Household Type (6)
    -   Unpaid Care (5)
