---
title: "Interactive Proportional Fitting"
subtitle: "Worked Example"
author: "Richard Blackwell"
format: 
  html:
    css: hisw.css
    grid:
      sidebar-width: 200px
      body-width: 1600px
      margin-width: 50px
      gutter-width: 1.5rem
---

```{r}
#| label: init
#| echo: false
#| warning: false
#| include: false

library(tidyverse)
library(readxl)
library(flextable)
library(officer)
```
## Data

The data we will use for the worked example comes from the UK Census 2021. 

We will use 3 marginals; **age**, **gender**, and **general health** for 4 selected census output areas (OA) 2021 in North Devon Local Authority District. They are

+ [E00101824] - Fremington
+ [E00101827] - Croyde
+ [E00165923] - Forches Avenue, Barnstaple
+ [E00180270] - Lankey

We will use 3 pairwise constraints; **age and gender**, **age and health**, and **gender and health** for the North Devon Local Authority District.

### Marginals

#### Age

The age marginal is given as 3 categories, young (0-14yrs), adult (16-64yrs), elderly (65+yrs) and the table below shows the age marginal for the 4 census output areas.

```{r}
#| label: tbl-age
#| tbl-cap: "Age"
#| echo: false
#| warning: false

df_age <- readxl::read_xlsx(path = "worked_example/data/marginals.xlsx",
                            sheet = "AGE") %>% 
  mutate(AREA = sprintf("[%s] - %s", OA21CD, OA21NM), 
         PCT_YOUNG = YOUNG/TOTAL, PCT_ADULT = ADULT/TOTAL, PCT_ELDERLY  = ELDERLY/TOTAL) %>%
  select(AREA, 
         YOUNG, ADULT, ELDERLY, TOTAL,
         PCT_YOUNG, PCT_ADULT, PCT_ELDERLY)

tbl_age <- flextable(df_age) %>%
  autofit() %>%
  add_header_row(values = c("Area", "n", "Proportion"), colwidths = c(1,4,3)) %>%
  align(i = 1, j = 2:8, align = "center", part = "header") %>%
  merge_at(j = 1, part = "header") %>%
  set_header_labels(AREA = "Area",
                    YOUNG = "Young",
                    ADULT = "Adult",
                    ELDERLY = "Elderly",
                    TOTAL = "Total",
                    PCT_YOUNG = "Young",
                    PCT_ADULT = "Adult",
                    PCT_ELDERLY = "Elderly") %>%
  set_formatter(PCT_YOUNG = function(x){sprintf("%.4f", x)},
                PCT_ADULT = function(x){sprintf("%.4f", x)},
                PCT_ELDERLY = function(x){sprintf("%.4f", x)}) %>%
  vline(j = c(1, 4, 5), part = "all")
tbl_age
```  

#### Gender

The gender marginal is given as 2 categories, female, and male and the table below shows the gender marginal for the 4 census output areas.

```{r}
#| label: tbl-gender
#| tbl-cap: "Gender"
#| echo: false
#| warning: false

df_gender <- readxl::read_xlsx(path = "worked_example/data/marginals.xlsx",
                               sheet = "GENDER") %>% 
  mutate(AREA = sprintf("[%s] - %s", OA21CD, OA21NM), 
         PCT_FEMALE = FEMALE/TOTAL, PCT_MALE = MALE/TOTAL) %>%
  select(AREA, 
         FEMALE, MALE, TOTAL,
         PCT_FEMALE, PCT_MALE)

tbl_gender <- flextable(df_gender) %>%
  autofit() %>%
  add_header_row(values = c("Area", "n", "Proportion"), colwidths = c(1,3,2)) %>%
  align(i = 1, j = 2:6, align = "center", part = "header") %>%
  merge_at(j = 1, part = "header") %>%
  set_header_labels(AREA = "Area",
                    FEMALE = "Female",
                    MALE = "Male",
                    TOTAL = "Total",
                    PCT_FEMALE = "Female",
                    PCT_MALE = "Male") %>%
  set_formatter(PCT_FEMALE = function(x){sprintf("%.4f", x)},
                PCT_MALE = function(x){sprintf("%.4f", x)}) %>%
  vline(j = c(1, 3, 4), part = "all")
tbl_gender
```

#### Health

The health marginal is given as 3 categories, good (very good and good health), fair (fair health), poor (bad and very bad health) and the table below shows the health marginal for the 4 census output areas.

```{r}
#| label: tbl-health
#| tbl-cap: "General Health"
#| echo: false
#| warning: false

df_health <- readxl::read_xlsx(path = "worked_example/data/marginals.xlsx",
                               sheet = "HEALTH") %>% 
  mutate(AREA = sprintf("[%s] - %s", OA21CD, OA21NM), 
         PCT_GOOD = GOOD/TOTAL, PCT_FAIR = FAIR/TOTAL, PCT_POOR = POOR/TOTAL) %>%
  select(AREA, 
         GOOD, FAIR, POOR, TOTAL,
         PCT_GOOD, PCT_FAIR, PCT_POOR)

tbl_health <- flextable(df_health) %>%
  autofit() %>%
  add_header_row(values = c("Area", "n", "Proportion"), colwidths = c(1,4,3)) %>%
  align(i = 1, j = 2:8, align = "center", part = "header") %>%
  merge_at(j = 1, part = "header") %>%
  set_header_labels(AREA = "Area",
                    GOOD = "Good",
                    FAIR = "Fair",
                    POOR = "Poor",
                    TOTAL = "Total",
                    PCT_GOOD = "Good",
                    PCT_FAIR = "Fair",
                    PCT_POOR = "Poor") %>%
  set_formatter(PCT_GOOD = function(x){sprintf("%.4f", x)},
                PCT_FAIR = function(x){sprintf("%.4f", x)},
                PCT_POOR = function(x){sprintf("%.4f", x)}) %>%
  vline(j = c(1, 4, 5), part = "all")
tbl_health
```

### Constraints

#### Age and Gender

The age (3 categories) and gender (2 categories) constraint based on the North Devon Local Authority District is give below.

```{r}
#| label: tbl-age-gender
#| tbl-cap: "Age and Gender"
#| echo: false
#| warning: false

df_age_gender <- readxl::read_xlsx(path = "worked_example/data/constraints.xlsx",
                                   sheet = "AGE_GENDER") %>%
  mutate(AREA = sprintf("[%s] - %s", LAD24CD, LAD24NM),
         PCT = POPN/sum(POPN)) %>%
  select(AREA, AGE, GENDER, POPN, PCT)

tbl_display <- df_age_gender %>%
  mutate(AGE = str_to_title(AGE),
         GENDER = str_to_title(GENDER)) %>%
  pivot_wider(names_from = "GENDER", values_from = c("POPN", "PCT")) %>%
  rename_with(.fn = ~c("AREA", "AGE", "FEMALE", "MALE", "PCT_FEMALE", "PCT_MALE")) %>%
  mutate(TOTAL = FEMALE + MALE, .after = "MALE") %>%
  mutate(PCT_TOTAL = TOTAL/sum(TOTAL), .after = "PCT_MALE") %>%
  bind_rows(
    df_age_gender %>% 
      pivot_wider(names_from = "GENDER", values_from = c("POPN", "PCT")) %>%
      mutate(TOTAL = POPN_FEMALE + POPN_MALE, .after = "POPN_MALE") %>%
      mutate(PCT_TOTAL = TOTAL/sum(TOTAL), .after = "PCT_MALE") %>%
      group_by(AREA) %>%
      summarise(across(.cols = 2:7, .fns = sum)) %>%
      ungroup() %>%
      mutate(AGE = "Total", .after = "AREA") %>%
      rename_with(.fn = ~c("AREA", "AGE", "FEMALE", "MALE", "TOTAL", "PCT_FEMALE", "PCT_MALE", "PCT_TOTAL")))
                              
tbl_age_gender <- flextable(tbl_display) %>%
  autofit() %>%
  add_header_row(values = c("Area", "Age", "n", "Proportion"), colwidths = c(1, 1, 3, 3)) %>%
  align(i = 1, j = 3:8, align = "center", part = "header") %>%
  set_header_labels(AREA = "Area",
                    AGE = "Age",
                    FEMALE = "Female",
                    MALE = "Male",
                    TOTAL = "Total",
                    PCT_FEMALE = "Female",
                    PCT_MALE = "Male",
                    PCT_TOTAL = "Total") %>%
  merge_v(j = 1:2, part = "header") %>%
  merge_at(j = 1, part = "body") %>%
  set_formatter(PCT_FEMALE = function(x){sprintf("%.4f", x)},
                PCT_MALE = function(x){sprintf("%.4f", x)},
                PCT_TOTAL = function(x){sprintf("%.4f", x)}) %>%
  vline(j = c(1, 2, 4:5, 7), part = "all") %>%
  hline(i = 3, part = "body")
tbl_age_gender
```

#### Age and Health

The age (3 categories) and health (3 categories) constraint based on the North Devon Local Authority District is give below.

```{r}
#| label: tbl-age-health
#| tbl-cap: "Age and Health"
#| echo: false
#| warning: false

df_age_health <- readxl::read_xlsx(path = "worked_example/data/constraints.xlsx",
                                   sheet = "AGE_HEALTH") %>%
  mutate(AREA = sprintf("[%s] - %s", LAD24CD, LAD24NM),
         PCT = POPN/sum(POPN)) %>%
  select(AREA, AGE, HEALTH, POPN, PCT)

tbl_display <- df_age_health %>%
  mutate(AGE = str_to_title(AGE),
         HEALTH = str_to_title(HEALTH)) %>%
  pivot_wider(names_from = "HEALTH", values_from = c("POPN", "PCT")) %>%
  rename_with(.fn = ~c("AREA", "AGE", "GOOD", "FAIR", "POOR", "PCT_GOOD", "PCT_FAIR", "PCT_POOR")) %>%
  mutate(TOTAL = GOOD + FAIR + POOR, .after = "POOR") %>%
  mutate(PCT_TOTAL = TOTAL/sum(TOTAL), .after = "PCT_POOR") %>%
  bind_rows(
    df_age_health %>% 
      pivot_wider(names_from = "HEALTH", values_from = c("POPN", "PCT")) %>%
      mutate(TOTAL = POPN_GOOD + POPN_FAIR + POPN_POOR, .after = "POPN_POOR") %>%
      mutate(PCT_TOTAL = TOTAL/sum(TOTAL), .after = "PCT_POOR") %>%
      group_by(AREA) %>%
      summarise(across(.cols = 2:9, .fns = sum)) %>%
      ungroup() %>%
      mutate(AGE = "Total", .after = "AREA") %>%
      rename_with(.fn = ~c("AREA", "AGE", "GOOD", "FAIR", "POOR", "TOTAL", "PCT_GOOD", "PCT_FAIR", "PCT_POOR", "PCT_TOTAL")))
                              
tbl_age_health <- flextable(tbl_display) %>%
  autofit() %>%
  add_header_row(values = c("Area", "Age", "n", "Proportion"), colwidths = c(1, 1, 4, 4)) %>%
  align(i = 1, j = 3:10, align = "center", part = "header") %>%
  set_header_labels(AREA = "Area",
                    AGE = "Age",
                    GOOD = "Good",
                    FAIR = "Fair",
                    POOR = "Poor",
                    TOTAL = "Total",
                    PCT_GOOD = "Good",
                    PCT_FAIR = "Fair",
                    PCT_POOR = "Poor",
                    PCT_TOTAL = "Total") %>%
  merge_v(j = 1:2, part = "header") %>%
  merge_at(j = 1, part = "body") %>%
  set_formatter(PCT_GOOD = function(x){sprintf("%.4f", x)},
                PCT_FAIR = function(x){sprintf("%.4f", x)},
                PCT_POOR = function(x){sprintf("%.4f", x)},
                PCT_TOTAL = function(x){sprintf("%.4f", x)}) %>%
  vline(j = c(1, 2, 5:6, 9), part = "all") %>%
  hline(i = 3, part = "body")
tbl_age_health
```


#### Gender and Health

The gender (2 categories) and health (3 categories) constraint based on the North Devon Local Authority District is give below.

```{r}
#| label: tbl-gender-health
#| tbl-cap: "Gender and Health"
#| echo: false
#| warning: false

df_gender_health <- readxl::read_xlsx(path = "worked_example/data/constraints.xlsx",
                                   sheet = "GENDER_HEALTH") %>%
  mutate(AREA = sprintf("[%s] - %s", LAD24CD, LAD24NM),
         PCT = POPN/sum(POPN)) %>%
  select(AREA, GENDER, HEALTH, POPN, PCT)

tbl_display <- df_gender_health %>%
  mutate(GENDER = str_to_title(GENDER),
         HEALTH = str_to_title(HEALTH)) %>%
  pivot_wider(names_from = "HEALTH", values_from = c("POPN", "PCT")) %>%
  rename_with(.fn = ~c("AREA", "GENDER", "GOOD", "FAIR", "POOR", "PCT_GOOD", "PCT_FAIR", "PCT_POOR")) %>%
  mutate(TOTAL = GOOD + FAIR + POOR, .after = "POOR") %>%
  mutate(PCT_TOTAL = TOTAL/sum(TOTAL), .after = "PCT_POOR") %>%
  bind_rows(
    df_gender_health %>% 
      pivot_wider(names_from = "HEALTH", values_from = c("POPN", "PCT")) %>%
      mutate(TOTAL = POPN_GOOD + POPN_FAIR + POPN_POOR, .after = "POPN_POOR") %>%
      mutate(PCT_TOTAL = TOTAL/sum(TOTAL), .after = "PCT_POOR") %>%
      group_by(AREA) %>%
      summarise(across(.cols = 2:9, .fns = sum)) %>%
      ungroup() %>%
      mutate(GENDER = "Total", .after = "AREA") %>%
      rename_with(.fn = ~c("AREA", "GENDER", "GOOD", "FAIR", "POOR", "TOTAL", "PCT_GOOD", "PCT_FAIR", "PCT_POOR", "PCT_TOTAL")))
                              
tbl_gender_health <- flextable(tbl_display) %>%
  autofit() %>%
  add_header_row(values = c("Area", "Gender", "n", "Proportion"), colwidths = c(1, 1, 4, 4)) %>%
  align(i = 1, j = 3:10, align = "center", part = "header") %>%
  set_header_labels(AREA = "Area",
                    GENDER = "Gender",
                    GOOD = "Good",
                    FAIR = "Fair",
                    POOR = "Poor",
                    TOTAL = "Total",
                    PCT_GOOD = "Good",
                    PCT_FAIR = "Fair",
                    PCT_POOR = "Poor",
                    PCT_TOTAL = "Total") %>%
  merge_v(j = 1:2, part = "header") %>%
  merge_at(j = 1, part = "body") %>%
  set_formatter(PCT_GOOD = function(x){sprintf("%.4f", x)},
                PCT_FAIR = function(x){sprintf("%.4f", x)},
                PCT_POOR = function(x){sprintf("%.4f", x)},
                PCT_TOTAL = function(x){sprintf("%.4f", x)}) %>%
  vline(j = c(1, 2, 5:6, 9), part = "all") %>%
  hline(i = 2, part = "body")
tbl_gender_health
```

## Iterative Proportional Fitting (IPF)

At a high level the process for **Iterative Proportional Fitting** is as follows

```{mermaid}
flowchart TD
    A(Seed Creation) --> B(IPF Balancing)
    B --> C{Convergence}
    C --> |Yes| D(End)
    C --> |No| B
```

## Higher Level Geography IPF

### Seed Creation

We will start by creating the overall seed table for the North Devon Local Authority District; age (3 categories), gender (2 categories), and health (3 categories). 

This seed table is a 3x2x3 table or in long format 18 rows as seen in @tbl-high-level-geography-seed-construction

The initial seed will be the mutual independent probability from the high level geography constraints. The assumption here is that each marginal calculated from the constraint table is the same **NB: this assumption should be checked as the census uses perturbation for privacy**

For example in @tbl-age-gender we can see that the gender split is Female 0.5137, Male 0.4863 and in @tbl-gender-health we can see that is also Female 0.5137, Male 0.4863

So from these marginals of the constraints we can create the following seed table

```{r}
#| label: tbl-high-level-geography-seed-construction
#| tbl-cap: "High Level Geography Seed Construction"
#| echo: false
#| warning: false

df_tmp <- df_age_gender %>% 
  group_by(AGE) %>% 
  summarise(POPN = sum(POPN)) %>% 
  ungroup() %>% 
  mutate(PCT = POPN/sum(POPN))

p_age <- setNames(object = df_tmp$PCT, nm = df_tmp$AGE)

df_tmp <- df_age_health %>% 
  group_by(HEALTH) %>% 
  summarise(POPN = sum(POPN)) %>% 
  ungroup() %>% 
  mutate(PCT = POPN/sum(POPN))

p_health <- setNames(object = df_tmp$PCT, nm = df_tmp$HEALTH)

df_tmp <- df_gender_health %>% 
  group_by(GENDER) %>% 
  summarise(POPN = sum(POPN)) %>% 
  ungroup() %>% 
  mutate(PCT = POPN/sum(POPN))

p_gender <- setNames(object = df_tmp$PCT, nm = df_tmp$GENDER)

df_seed <- data.frame(AREA = "[E07000043] - North Devon",
                      AGE = c(rep("YOUNG", 6), rep("ADULT", 6), rep("ELDERLY", 6)),
                      GENDER = rep(c(rep("FEMALE", 3), rep("MALE", 3)), 3),
                      HEALTH = rep(c("GOOD", "FAIR", "POOR"), 6)) %>%
  mutate(P_AGE = p_age[AGE], 
         P_GENDER = p_gender[GENDER], 
         P_HEALTH = p_health[HEALTH],
         P_OVERALL = P_AGE * P_GENDER * P_HEALTH)

age_categories <- setNames(c("Young", "Adult", "Elderly"), nm = c("YOUNG", "ADULT", "ELDERLY"))
gender_categories <- setNames(c("Female", "Male"), nm = c("FEMALE", "MALE"))
health_categories <- setNames(c("Good", "Fair", "Poor"), nm = c("GOOD", "FAIR", "POOR"))

tbl_seed <- flextable(df_seed %>% 
                        mutate(AGE = age_categories[AGE],
                               GENDER = gender_categories[GENDER],
                               HEALTH = health_categories[HEALTH])) %>%
  autofit() %>%
  add_header_row(values = c("Area", "Category", "Probability"), colwidths = c(1, 3, 4)) %>%
  align(i = 1, j = c(2:4, 5:8), align = "center", part = "header") %>%
  set_header_labels(AREA = "Area",
                    AGE = "Age",
                    GENDER = "Gender",
                    HEALTH = "Health",
                    P_AGE = "P(Age)",
                    P_GENDER = "P(Gender)",
                    P_HEALTH = "P(Health)",
                    P_OVERALL = "P(Age,Gender,Health)") %>%
  set_formatter(P_AGE = function(x){sprintf("%.4f", x)},
                P_GENDER = function(x){sprintf("%.4f", x)},
                P_HEALTH = function(x){sprintf("%.4f", x)},
                P_OVERALL = function(x){sprintf("%.4f", x)}) %>%
  merge_v(j = 1:4) %>%
  vline(j = c(1, 4, 7), border = fp_border(color = "grey80")) %>%
  hline(i = 1:17, border = fp_border(color = "grey80"))
tbl_seed
```

### IPF Balancing

In this step we will adjust the seed table to balance to, in the case of the high level geography, the pairwise constraints

In order to do this we will step through each of the 3 constraints and balance them in turn.

#### Age and Gender Constraint

We begin by calculating the **age and gender** table based on the seed table by summing each of the age and gender combinations. For example the **young, female** proportion is the sum of the **young, female, good health**, the **young, female, fair health**, and the **young, female, poor health** proportions.

----

$$
  \begin{aligned}
    P(age, gender) &= \sum_{health}P(age, gender, health) \\
    P(young, female) &= P(young, female, good) + P(young, female, fair) + P(young, female, poor) \\
    P(young, female) &= 0.0700 + 0.0119 + 0.0047 \\
    P(young, female) &= 0.0866 \\
  \end{aligned}
$$

----

```{r}
#| label: tbl-high-level-geography-age-gender-balance-step1
#| tbl-cap: "High Level Geography Age Gender Balance (Step 1)"
#| echo: false
#| warning: false

df_age_gender_balance <- df_age_gender %>% 
  mutate(AGE, GENDER, P_ORIGINAL = PCT, .keep = "none") %>%
  left_join(df_seed %>% 
              group_by(AGE, GENDER) %>% 
              summarise(P_0 = sum(P_OVERALL), .groups = "keep") %>% 
              ungroup(),
            by = c("AGE", "GENDER")) %>%
  mutate(ADJ_0 = P_ORIGINAL / P_0)

tbl_age_gender_balance <- flextable(df_age_gender_balance %>%
                                      mutate(AGE = age_categories[AGE],
                                             GENDER = gender_categories[GENDER])) %>% 
  autofit() %>%
  add_header_row(values = c("Age", "Gender", "P(Age, Gender)", "Adjustment Factor\n(Orig/Calc)"),
                 colwidths = c(1,1,2,1)) %>%
  set_header_labels(AGE = "Age",
                    GENDER = "Gender",
                    P_ORIGINAL = "Constraint",
                    P_0 = "Seed",
                    ADJ_0 = "Adjustment Factor\n(Orig/Calc)") %>%
  merge_v(part = "header") %>%
  set_formatter(P_ORIGINAL = function(x){sprintf("%.4f", x)},
                P_0 = function(x){sprintf("%.4f", x)},
                ADJ_0 = function(x){sprintf("%.4f", x)}) %>%
  vline(j = c(2, 4))
tbl_age_gender_balance
```

We then adjust all the proportions in the seed table by the relevant adjustment factor, for example the **young, female, good health**, **young, female, fair health**, and **young, female, poor health** will be multiplied by the adjustment factor **0.9594** which results in the following table.

```{r}
#| label: tbl-high-level-geography-age-gender-iter1
#| tbl-cap: "High Level Geography Age Gender Constraint Balancing Iteration 1"
#| echo: false
#| warning: false

df_seed_iter1 <- df_seed %>% 
  mutate(AGE, GENDER, HEALTH, P_SEED = P_OVERALL, .keep = "none") %>%
  left_join(df_age_gender_balance %>% mutate(AGE, GENDER, ADJ_0_AG = ADJ_0, .keep = "none"), by = c("AGE", "GENDER")) %>%
  mutate(P_0_AG = P_SEED * ADJ_0_AG)

tbl_iter1 <- flextable(df_seed_iter1 %>% 
                        mutate(AGE = age_categories[AGE],
                               GENDER = gender_categories[GENDER],
                               HEALTH = health_categories[HEALTH])) %>%
  autofit() %>%
  add_header_row(values = c("Age", "Gender", "Health", "Seed", "Iteration 1 (Age, Gender)"), colwidths = c(1, 1, 1, 1, 2)) %>%
  align(i = 1, j = c(5:6), align = "center", part = "header") %>%
  set_header_labels(AGE = "Age",
                    GENDER = "Gender",
                    HEALTH = "Health",
                    P_SEED = "Seed",
                    ADJ_0_AG = "Adjustment Factor",
                    P_0_AG = "P Adjusted") %>%
  set_formatter(P_SEED = function(x){sprintf("%.4f", x)},
                ADJ_0_AG = function(x){sprintf("%.4f", x)},
                P_0_AG = function(x){sprintf("%.4f", x)}) %>%
  merge_v(j = 1:4, part = "header") %>%
  merge_v(j = 1:3) %>%
  vline(j = c(3, 4), border = fp_border(color = "grey80")) %>%
  hline(i = 1:17, border = fp_border(color = "grey80"))
tbl_iter1
```

After the balancing we can see that age and gender combinations are now balanced to the high level geography age and gender constraint. 

For example the **young, female** proportion is the sum of the **young, female, good health** plus the **young, female, fair health** plus the **young, female, poor health** proportions.

----

$$
  \begin{aligned}
    P(age, gender) &= \sum_{health}P(age, gender, health) \\
    P(young, female) &= P(young, female, good) + P(young, female, fair) + P(young, female, poor) \\
    P(young, female) &= 0.0672 + 0.0114 + 0.0045 \\
    P(young, female) &= 0.0831 \\
  \end{aligned}
$$

----

Which we can see from @tbl-age-gender is the same as the original constraint value.

#### Age and Health Constraint

We now move on to balancing by the next constraint which is age and health. In the same process as before we calculate the current proportions by summing each of the **age and health** combinations that we have just created by balancing for **age and gender**. For example the **young, good** proportion is the sum of the **young, female, good health** plus the **young, male, good health** proportions.

----

$$
  \begin{aligned}
    P(age, health) &= \sum_{gender}P(age, gender, health) \\
    P(young, good) &= P(young, female, good) + P(young, male, good) \\
    P(young, good) &= 0.0672 + 0.0692 \\
    P(young, good) &= 0.1364 \\
  \end{aligned}
$$

----

```{r}
#| label: tbl-high-level-geography-age-health-balance-step1
#| tbl-cap: "High Level Geography Age Health Balance (Step 1)"
#| echo: false
#| warning: false

df_age_health_balance <- df_age_health %>% 
  mutate(AGE, HEALTH, P_ORIGINAL = PCT, .keep = "none") %>%
  left_join(df_seed_iter1 %>% 
              group_by(AGE, HEALTH) %>% 
              summarise(P_0 = sum(P_0_AG), .groups = "keep") %>% 
              ungroup(),
            by = c("AGE", "HEALTH")) %>%
  mutate(ADJ_0 = P_ORIGINAL / P_0)

tbl_age_health_balance <- flextable(df_age_health_balance %>%
                                      mutate(AGE = age_categories[AGE],
                                             HEALTH = health_categories[HEALTH])) %>% 
  autofit() %>%
  add_header_row(values = c("Age", "Health", "P(Age, Health)", "Adjustment Factor\n(Orig/Calc)"),
                 colwidths = c(1,1,2,1)) %>%
  set_header_labels(AGE = "Age",
                    HEALTH = "Health",
                    P_ORIGINAL = "Constraint",
                    P_0 = "Seed",
                    ADJ_0 = "Adjustment Factor\n(Orig/Calc)") %>%
  merge_v(part = "header") %>%
  set_formatter(P_ORIGINAL = function(x){sprintf("%.4f", x)},
                P_0 = function(x){sprintf("%.4f", x)},
                ADJ_0 = function(x){sprintf("%.4f", x)}) %>%
  vline(j = c(2, 4))
tbl_age_health_balance
```

As before We adjust all the proportions in the seed table by the relevant adjustment factor, for example the **young, female, good health** and **young, male, good health** will be multiplied by the adjustment factor **1.2041** which results in the following table.

```{r}
#| label: tbl-high-level-geography-age-health-iter1
#| tbl-cap: "High Level Geography Age Health Constraint Balancing Iteration 1"
#| echo: false
#| warning: false

df_seed_iter1 <- df_seed_iter1 %>% 
  left_join(df_age_health_balance %>% mutate(AGE, HEALTH, ADJ_0_AH = ADJ_0, .keep = "none"), by = c("AGE", "HEALTH")) %>%
  mutate(P_0_AH = P_0_AG * ADJ_0_AH)

tbl_iter1 <- flextable(df_seed_iter1 %>% 
                        mutate(AGE = age_categories[AGE],
                               GENDER = gender_categories[GENDER],
                               HEALTH = health_categories[HEALTH])) %>%
  autofit() %>%
  add_header_row(values = c("Age", "Gender", "Health", "Seed", "Iteration 1 (Age, Gender)", "Iteration 1 (Age, Health)"), colwidths = c(1, 1, 1, 1, 2, 2)) %>%
  align(i = 1, j = c(5:8), align = "center", part = "header") %>%
  set_header_labels(AGE = "Age",
                    GENDER = "Gender",
                    HEALTH = "Health",
                    P_SEED = "Seed",
                    ADJ_0_AG = "Adjustment Factor",
                    P_0_AG = "P Adjusted",
                    ADJ_0_AH = "Adjustment Factor",
                    P_0_AH = "P Adjusted") %>%
  set_formatter(P_SEED = function(x){sprintf("%.4f", x)},
                ADJ_0_AG = function(x){sprintf("%.4f", x)},
                P_0_AG = function(x){sprintf("%.4f", x)},
                ADJ_0_AH = function(x){sprintf("%.4f", x)},
                P_0_AH = function(x){sprintf("%.4f", x)}) %>%
  merge_v(j = 1:8, part = "header") %>%
  merge_v(j = 1:3) %>%
  vline(j = c(3, 4, 6), border = fp_border(color = "grey80")) %>%
  hline(i = 1:17, border = fp_border(color = "grey80"))
tbl_iter1
```

After the balancing we can see that age and health combinations are now balanced to the high level geography age and health constraint. 

For example the **young, good** proportion is the sum of the **young, female, good health** and the **young, male, good health** proportions.

----

$$
  \begin{aligned}
    P(age, health) &= \sum_{gender}P(age, gender, health) \\
    P(young, good) &= P(young, female, good) + P(young, male, good) \\
    P(young, good) &= 0.0809 + 0.0833 \\
    P(young, good) &= 0.1642 \\
  \end{aligned}
$$

----

Which we can see from @tbl-age-health is the same as the original constraint value.

#### Gender and Health Constraint

We now move on to balancing by the final constraint which is **gender and health** in the same way as we did with the other tow constraints. We calculate the current proportions by summing each of the **gender and health** combinations that we have just created by balancing for **age and health**. For example the **female, good** proportion is the sum of the **young, female, good health**, the **adult, female, good health**, and **elderly, female, good health**  proportions.

----

$$
  \begin{aligned}
    P(gender, health) &= \sum_{age}P(age, gender, health) \\
    P(female, good) &= P(young, female, good) + P(adult, female, good) + P(adult, female, good) \\
    P(female, good) &= 0.0809 + 0.2491 + 0.0838 \\
    P(female, good) &= 0.4138 \\
  \end{aligned}
$$

----

```{r}
#| label: tbl-high-level-geography-gender-health-balance-step1
#| tbl-cap: "High Level Geography Gender Health Balance (Step 1)"
#| echo: false
#| warning: false

df_gender_health_balance <- df_gender_health %>% 
  mutate(GENDER, HEALTH, P_ORIGINAL = PCT, .keep = "none") %>%
  left_join(df_seed_iter1 %>% 
              group_by(GENDER, HEALTH) %>% 
              summarise(P_0 = sum(P_0_AH), .groups = "keep") %>% 
              ungroup(),
            by = c("GENDER", "HEALTH")) %>%
  mutate(ADJ_0 = P_ORIGINAL / P_0)

tbl_gender_health_balance <- flextable(df_gender_health_balance %>%
                                         mutate(GENDER = gender_categories[GENDER],
                                                HEALTH = health_categories[HEALTH])) %>% 
  autofit() %>%
  add_header_row(values = c("Gender", "Health", "P(Gender, Health)", "Adjustment Factor\n(Orig/Calc)"),
                 colwidths = c(1,1,2,1)) %>%
  set_header_labels(GENDER = "Gender",
                    HEALTH = "Health",
                    P_ORIGINAL = "Constraint",
                    P_0 = "Seed",
                    ADJ_0 = "Adjustment Factor\n(Orig/Calc)") %>%
  merge_v(part = "header") %>%
  set_formatter(P_ORIGINAL = function(x){sprintf("%.4f", x)},
                P_0 = function(x){sprintf("%.4f", x)},
                ADJ_0 = function(x){sprintf("%.4f", x)}) %>%
  vline(j = c(2, 4))
tbl_gender_health_balance
```

As before We adjust all the proportions in the seed table by the relevant adjustment factor, for example the **young, female, good health**, **adult, female, good health**, and **elderly, female, good health** will be multiplied by the adjustment factor **0.9990** which results in the following table.

```{r}
#| label: tbl-high-level-geography-gender-health-iter1
#| tbl-cap: "High Level Geography Gender Health Constraint Balancing Iteration 1"
#| echo: false
#| warning: false

df_seed_iter1 <- df_seed_iter1 %>% 
  left_join(df_gender_health_balance %>% mutate(GENDER, HEALTH, ADJ_0_GH = ADJ_0, .keep = "none"), by = c("GENDER", "HEALTH")) %>%
  mutate(P_0_GH = P_0_AH * ADJ_0_GH)

tbl_iter1 <- flextable(df_seed_iter1 %>% 
                        mutate(AGE = age_categories[AGE],
                               GENDER = gender_categories[GENDER],
                               HEALTH = health_categories[HEALTH])) %>%
  autofit() %>%
  add_header_row(values = c("Age", "Gender", "Health", "Seed", "Iteration 1 (Age, Gender)", "Iteration 1 (Age, Health)", "Iteration 1 (Gender, Health)"), colwidths = c(1, 1, 1, 1, 2, 2, 2)) %>%
  align(i = 1, j = c(5:10), align = "center", part = "header") %>%
  set_header_labels(AGE = "Age",
                    GENDER = "Gender",
                    HEALTH = "Health",
                    P_SEED = "Seed",
                    ADJ_0_AG = "Adjustment Factor",
                    P_0_AG = "P Adjusted",
                    ADJ_0_AH = "Adjustment Factor",
                    P_0_AH = "P Adjusted",
                    ADJ_0_GH = "Adjustment Factor",
                    P_0_GH = "P Adjusted") %>%
  set_formatter(P_SEED = function(x){sprintf("%.4f", x)},
                ADJ_0_AG = function(x){sprintf("%.4f", x)},
                P_0_AG = function(x){sprintf("%.4f", x)},
                ADJ_0_AH = function(x){sprintf("%.4f", x)},
                P_0_AH = function(x){sprintf("%.4f", x)},
                ADJ_0_GH = function(x){sprintf("%.4f", x)},
                P_0_GH = function(x){sprintf("%.4f", x)}) %>%
  merge_v(j = 1:10, part = "header") %>%
  merge_v(j = 1:3) %>%
  vline(j = c(3, 4, 6, 8), border = fp_border(color = "grey80")) %>%
  hline(i = 1:17, border = fp_border(color = "grey80"))
tbl_iter1
```

After the balancing we can see that gender and health combinations are now balanced to the high level geography gender and health constraint. 

For example the **female, good** proportion is the sum of the **young, female, good health**, **adult, female, good health**, and **elderly, female, good health** proportions.

----

$$
  \begin{aligned}
    P(gender, health) &= \sum_{gender}P(age, gender, health) \\
    P(female, good) &= P(young, female, good) + P(elderly, female, good) + P(elderly, female, good) \\
    P(female, good) &= 0.0808 + 0.2489 + 0.0838 \\
    P(female, good) &= 0.4135 \\
  \end{aligned}
$$

----

Which we can see from @tbl-gender-health is the same as the original constraint value (NB: 0.4135 vs 0.4134 due to rounding effects).

### Covergence 

Each time we balanced for one of the constraints that constraint would be balance to the high level geography but that in turn would unbalance the other constraints.

So if we look at the **gender and health** constraint the summed categories from the latest iteration match the original high level gender health constraint as this was the most recent constraint that we balanced.

```{r}
#| label: tbl-high-level-geography-gender-health-diff
#| tbl-cap: "High Level Geography Gender Health Difference Iteration 1"
#| echo: false
#| warning: false

df_gender_health_diff <- df_gender_health %>% mutate(GENDER, HEALTH, P_ORIG = PCT, .keep = "none") %>%
  left_join(df_seed_iter1 %>% group_by(GENDER, HEALTH) %>% summarise(P1 = sum(P_0_GH), .groups = "keep") %>% ungroup(),
            by = c("GENDER", "HEALTH")) %>%
  mutate(ABS_DIFF = abs(P_ORIG - P1)) %>% 
  arrange(GENDER, HEALTH)

tbl_gender_health_diff <- flextable(df_gender_health_diff %>%
                                      mutate(GENDER = gender_categories[GENDER],
                                             HEALTH = health_categories[HEALTH])) %>%
  autofit() %>%
  add_header_row(values = c("Gender", "Health", "Constraint" , "Iteration 1"), colwidths = c(1, 1, 1, 2)) %>%
  align(i = 1, j = 4:5, align = "center", part = "header") %>%
  set_header_labels(GENDER = "Gender",
                    HEALTH = "Health",
                    P_ORIG = "Constraint",
                    P1 = "Iteration 1",
                    ABS_DIFF = "Abs. Diff.") %>%
  merge_v(j = 1:3, part = "header") %>%
  merge_v(j = 1:2, part = "body") %>%
  set_formatter(P_ORIG = function(x){sprintf("%.9f", x)},
                P1 = function(x){sprintf("%.9f", x)},
                ABS_DIFF = function(x){sprintf("%.9f", x)}) %>%
  vline(j = c(2, 3), border = fp_border(color = "grey80")) %>%
  hline(i = 1:5, border = fp_border(color = "grey80"))
tbl_gender_health_diff
```

However if we look at the **age and gender** constraint, the first we balanced. We can see that it has become unbalanced as there are differences between the original constraint and the sum of the categories after balancing for **gender and health**

```{r}
#| label: tbl-high-level-geography-age-gender-diff
#| tbl-cap: "High Level Geography Age Gender Difference Iteration 1"
#| echo: false
#| warning: false

df_age_gender_diff <- df_age_gender %>% mutate(AGE, GENDER, P_ORIG = PCT, .keep = "none") %>%
  left_join(df_seed_iter1 %>% group_by(AGE, GENDER) %>% summarise(P1 = sum(P_0_GH), .groups = "keep") %>% ungroup(),
            by = c("AGE", "GENDER")) %>%
  mutate(ABS_DIFF = abs(P_ORIG - P1)) %>% 
  arrange(AGE, GENDER)

tbl_age_gender_diff <- flextable(df_age_gender_diff %>%
                                   mutate(AGE = age_categories[AGE],
                                          GENDER = gender_categories[GENDER])) %>%
  autofit() %>%
  add_header_row(values = c("Age", "Gender", "Constraint" , "Iteration 1"), colwidths = c(1, 1, 1, 2)) %>%
  align(i = 1, j = 4:5, align = "center", part = "header") %>%
  set_header_labels(AGE = "Age",
                    GENDER = "Gender",
                    P_ORIG = "Constraint",
                    P1 = "Iteration 1",
                    ABS_DIFF = "Abs. Diff.") %>%
  merge_v(j = 1:3, part = "header") %>%
  merge_v(j = 1:2, part = "body") %>%
  set_formatter(P_ORIG = function(x){sprintf("%.9f", x)},
                P1 = function(x){sprintf("%.9f", x)},
                ABS_DIFF = function(x){sprintf("%.9f", x)}) %>%
  vline(j = c(2, 3), border = fp_border(color = "grey80")) %>%
  hline(i = 1:5, border = fp_border(color = "grey80"))
tbl_age_gender_diff
```

And if we do the same for the **age and health** constraint, we can again see that it has become unbalanced as there are differences between the original constraint and the sum of the categories after balancing for **gender and health**

```{r}
#| label: tbl-high-level-geography-age-health-diff
#| tbl-cap: "High Level Geography Age Health Difference Iteration 1"
#| echo: false
#| warning: false

df_age_health_diff <- df_age_health %>% mutate(AGE, HEALTH, P_ORIG = PCT, .keep = "none") %>%
  left_join(df_seed_iter1 %>% group_by(AGE, HEALTH) %>% summarise(P1 = sum(P_0_GH), .groups = "keep") %>% ungroup(),
            by = c("AGE", "HEALTH")) %>%
  mutate(ABS_DIFF = abs(P_ORIG - P1)) %>% 
  arrange(AGE, HEALTH)

tbl_age_health_diff <- flextable(df_age_health_diff %>%
                                   mutate(AGE = age_categories[AGE],
                                          HEALTH = health_categories[HEALTH])) %>%
  autofit() %>%
  add_header_row(values = c("Age", "Health", "Constraint" , "Iteration 1"), colwidths = c(1, 1, 1, 2)) %>%
  align(i = 1, j = 4:5, align = "center", part = "header") %>%
  set_header_labels(AGE = "Age",
                    HEALTH = "Health",
                    P_ORIG = "Constraint",
                    P1 = "Iteration 1",
                    ABS_DIFF = "Abs. Diff.") %>%
  merge_v(j = 1:3, part = "header") %>%
  merge_v(j = 1:2, part = "body") %>%
  set_formatter(P_ORIG = function(x){sprintf("%.9f", x)},
                P1 = function(x){sprintf("%.9f", x)},
                ABS_DIFF = function(x){sprintf("%.9f", x)}) %>%
  vline(j = c(2, 3), border = fp_border(color = "grey80")) %>%
  hline(i = 1:5, border = fp_border(color = "grey80"))
tbl_age_health_diff
```

From these 3 tables we can calculate the maximum error which is this instance is `r max(df_age_gender_diff$ABS_DIFF, df_age_health_diff$ABS_DIFF, df_gender_health_diff$ABS_DIFF)`