---
title: "Intersectionality using Iterative Proportional Fitting of 2021 UK Census Data"
author: "Richard Blackwell"
format: html
editor: visual
---

```{r}
#| label: init
#| echo: false
#| warning: false

library(tidyverse)
library(flextable)
library(officer)

```

## Worked Example

### Marginals

```{r}
#| label: marginal_popn
#| echo: false
#| warning: false

df_popn <- data.frame(area = c("A","B","C"),
                      popn = c(293, 222, 172))

flextable(df_popn) %>%
  autofit() %>%
  set_header_labels(area = "Area", popn = "Popn") %>%
  vline(j = 1, border = fp_border(width = 2))
```

```{r}
#| label: marginal_age
#| echo: false
#| warning: false

df_age <- data.frame(age = rep(c("Young", "Adult", "Elderly"), 3),
                     area = c(rep("A", 3), rep("B", 3), rep("C", 3)),
                     popn = c(c(90, 133, 70), c(62, 107, 53), c(45, 83, 44)))
df_age <- df_age %>% group_by(area) %>% mutate(pct = popn/sum(popn)) %>% ungroup() %>% select(-popn)

flextable(df_age %>% pivot_wider(names_from = "area", values_from = "pct")) %>%
  autofit() %>%
  align(j = 2:4, align = "center", part = "header") %>%
  add_header_row(values = c("Age", "Area"), colwidths = c(1, 3)) %>%
  set_header_labels(age = "Age") %>%
  merge_at(i = 1:2, j = 1, part = "header") %>%
  vline(j = 1, border = fp_border(width = 2)) %>%
  colformat_double(j = 1:length(unique(df_age$area))+1, digits=3)
```

```{r}
#| label: marginal_gender
#| echo: false
#| warning: false

df_gender <- data.frame(gender = rep(c("Female", "Male"), 3),
                        area = c(rep("A", 2), rep("B", 2), rep("C", 2)),
                        popn = c(c(153, 140), c(116, 106), c(90, 82)))

df_gender <- df_gender %>% group_by(area) %>% mutate(pct = popn/sum(popn)) %>% ungroup() %>% select(-popn)

flextable(df_gender %>% pivot_wider(names_from = "area", values_from = "pct")) %>%
  autofit() %>%
  align(j = 2:4, align = "center", part = "header") %>%
  add_header_row(values = c("Gender", "Area"), colwidths = c(1, 3)) %>%
  set_header_labels(age = "Gender") %>%
  merge_at(i = 1:2, j = 1, part = "header") %>%
  vline(j = 1, border = fp_border(width = 2)) %>%
  colformat_double(j = 1:length(unique(df_gender$area))+1, digits=3)
```

```{r}
#| label: marginal_health
#| echo: false
#| warning: false

df_health <- data.frame(health = rep(c("Good", "Fair", "Poor"), 3),
                        area = c(rep("A", 3), rep("B", 3), rep("C", 3)),
                        popn = c(c(168, 82, 43), c(127, 63, 32), c(96, 51, 25)))

df_health <- df_health %>% group_by(area) %>% mutate(pct = popn/sum(popn)) %>% ungroup() %>% select(-popn)

flextable(df_health %>% pivot_wider(names_from = "area", values_from = "pct")) %>%
  autofit() %>%
  align(j = 2:4, align = "center", part = "header") %>%
  add_header_row(values = c("Health", "Area"), colwidths = c(1, 3)) %>%
  set_header_labels(age = "Health") %>%
  merge_at(i = 1:2, j = 1, part = "header") %>%
  vline(j = 1, border = fp_border(width = 2)) %>%
  colformat_double(j = 1:length(unique(df_health$area))+1, digits=3)
```

### Constraints

```{r}
#| label: constraint_age_gender
#| echo: false
#| warning: false

df_age_gender <- data.frame(age = rep(c("Young", "Adult", "Elderly"), 2),
                            gender = c(rep("Female", 3), rep("Male", 3)),
                            popn = c(c(103, 163, 93), c(94, 160, 74)))
df_age_gender <- df_age_gender %>% mutate(pct = popn/sum(popn)) %>% select(-popn)
```

```{r}
#| label: constraint_age_health
#| echo: false
#| warning: false

df_age_health <- data.frame(age = rep(c("Young", "Adult", "Elderly"), 3),
                            health = c(rep("Good", 3), rep("Fair", 3), rep("Poor", 3)),
                            popn = c(c(120, 187, 84), c(48, 94, 54), c(29, 42, 29)))
df_age_health <- df_age_health %>% mutate(pct = popn/sum(popn)) %>% select(-popn)
```

```{r}
#| label: constraint_gender_health
#| echo: false
#| warning: false

df_gender_health <- data.frame(gender = rep(c("Female", "Male"), 3),
                               health = c(rep("Good", 2), rep("Fair", 2), rep("Poor", 2)),
                               popn = c(c(197, 194), c(104, 92), c(58, 42)))
df_gender_health <- df_gender_health %>% mutate(pct = popn/sum(popn)) %>% select(-popn)
```

### Seed Table

The starting seed table $X^{(0)}$ is given by the combined independent probability multiplied by the population to return the initial estimate of people with the combined values.

For example the seed estimate $X^{(0)}_{a,g,h}$ for young, male in good health is given by $$ \begin{align} X^{(0)}_{Y,M,G} &= P(Y)\times{P(M)}\times{P(G)} \\ &= 0.307\times{0.478}\times{0.573} \\ &= 0.084 \end{align} $$

Each cell of the seed table is calculated in the same manner to cover the $ age\times{gender}\times{health} = 3\times{2}\times{3} = 18 $ combinations resulting in the seed table as follows

```{r}
#| label: seed_table
#| echo: false
#| warning: false

area_list <- unique(df_age$area)
age_list <- unique(df_age$age)
gender_list <- unique(df_gender$gender)
health_list <- unique(df_health$health)

df_seed_table <- data.frame(area = as.character(),
                            age = as.character(),
                            gender = as.character(),
                            health = as.character(),
                            pct = as.numeric())

# seed_table <- array(dim = c(length(area_list), 
#               length(age_list),
#               length(gender_list),
#               length(health_list)),
#               dimnames = list(area_list, age_list, gender_list, health_list))

for(area in area_list){
  for(age in age_list){
    for(gender in gender_list){
      for(health in health_list){
        pct <- df_age$pct[df_age$area==area & df_age$age==age] *
          df_gender$pct[df_gender$area==area & df_gender$gender==gender] *
          df_health$pct[df_health$area==area & df_health$health==health]
        df_seed_table <- df_seed_table %>% bind_rows(data.frame(area, age, gender, health, pct))
      }
    }
  }
}

flextable(df_seed_table) %>%
  set_header_labels(area = "Area",
                    age = "Age",
                    gender = "Gender",
                    health = "Health",
                    pct = "Proportion") %>%
  bold(i = 1, part = "header") %>%
  border_inner(border = fp_border(width = 1)) %>%
  merge_v() %>%
  set_formatter(pct = function(x){sprintf("%.3f", x)})
```

### Create calculated constraints

```{r} 
#| label: calculated_constraint_age_gender
#| echo: false
#| warning: false

df_seed_table %>%
  group_by(age, gender) %>% 
  summarise(pct = sum(pct), .groups = "keep")

sum(df_seed_table$pct)
```